Unshuffling Permutations

### 2015-09-17 article.tex
Modifs intro (je continue ce soir)

Given three words $u$, $v$ and $w$, $u$ is said to a \emph{shuffle} of
$v$ and $w$ if it can be formed by interleaving the letters from $v$ and
$w$ in a way that maintains the left-to-right ordering of the letters
from each word (it is a classical textbook exercise to design an
efficient dynamic programming algorithm that determines whether $u$ is a
shuffle of $v$ and $w$). A word $u$ is said to be a \emph{square} (w.r.t.
the shuffle) if there exists a word $v$ such that $u$ is a shuffle of $v$
with itself.
\smallskip

This paper is intended to study a natural generalization of the shuffle
to permutations. Given permutations $\pi$, $\mu$ and $\nu$, $\pi$ is
said to be a shuffle of $\mu$ and $\nu$ if $\pi$ (viewed as a word) is
a shuffle of two words that are order-isomorphic to $\mu$ and $\nu$. For
example, $\pi$ is the shuffle of $\mu=$ and $\nu=$ as $\pi$. A
permutation $\pi$ is said to be a square (w.r.t. the shuffle) if there
exists a permutation $\mu$ such that $\pi$ is a shuffle of $\mu$ with
itself. For example, $\pi = 241356$ is a square as it is the shuffle of
$213$ with itself (\emph{e.g.},
$\pi = \dot{2}{4}\dot{1}{3}{5}\dot{6}$, where the $\dot{x}$'s
denote the second occurrence), whereas $\pi = 4123$ is not.
The shuffle product defined and considered in this paper is different
from other existing and well-known shuffle products on permutations.
For instance, in~\cite{DHT:IJAC:2002}, the authors defined the
{\em shifted shuffle product}. In this product $\pi$ is a shuffle of
$\mu$ and $\nu$ if $\pi$ is the shuffle, as words, of $\mu$ and the
word obtained by incrementing all letters of $\nu$ by the size of $\mu$.
It is a simple exercise to prove that, give a permutation $\pi$, the
decision problem to test if $\pi$ is a square for the shifted shuffle
product is linear.
\smallskip

Formally, the shuffle of two words $u$ and $v$ of $A^*$ is the language
$u \shuffle v$ consisting of all words $u_1 v_1 u_2 v_2 \ldots u_k v_k$,
where $k \geq 0$ and the $u_i$ and the $v_i$ are the words  of $A^*$
such that $u = u_1 u_2 \ldots u_k$ and $v = v_1 v_2 \ldots v_k$. We
first briefly review the key results for words that arise in our context.
Given words $u$, $v_1$ and $v_2$, it can be tested in
$O\left(|u|^2 / \log(|u|)\right)$ time whether or not
$u \in v_1 \shuffle v_2$~\cite{Leeuwen:Nivat:IPL:1982}.
(To the best of our knowledge, the first $O(|u|^2)$ time algorithm a
ppeared in~\cite{Mansfield:DAM:1983}. This algorithm can easily be
extended to check in polynomial-time whether or not a word is the
shuffle of any fixed number of given words.) The shuffle $u \shuffle v$
of words $u$ and $v$ can be computed in
$O\left((|u|+|v|) \; \binom{|u|+|v|}{|u|}\right)$ time~\cite{Spehner:TCS:1986}.
An improvement and generalization has been proposed in
\cite{Allauzen:IGM:2000}, where it is proved that, given words
$u_1, u_2, \ldots, u_k$, the shuffle $\shuffle_{i=1}^{k} u_i$ can be
computed in
$O\left(\binom{|u_1|+|u_2|+\ldots+|u_k|}{|u_1|,|u_2|,\ldots,|u_k|}\right)$
time. Given words $u, v_1, v_2, \ldots, v_n \in A^*$, it is however \NPC
to decide whether or not $u \in \shuffle_{i=1}^{k} v_i$.
\cite{Mansfield:DAM:1983,Warmuth:Haussler:JCSS:1984}.
This remains true even if the alphabet has size $3$~\cite{Warmuth:Haussler:JCSS:1984}.
Of particular interest,
it is shown in~\cite{Warmuth:Haussler:JCSS:1984} that this problem remains
\NPC even if all
words $v_1, v_2, \ldots, v_n$ are identical,
thereby proving that,
for two words $u$ and $v$, it is \NPC to decide whether or not $u$
is in the iterated shuffle of $v$.
Again, this remains true even if the alphabet has size $3$.
As for squares, given a word $z$ it is \NPC to determine whether
there exists a word $x$ such that $z$ is the shuffle of $x$ with itself
\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013}.
\smallskip

@@

We denote by $S_n$ the set of permutations of size $n$ and by $S$ the
set of all permutations. In this paper, permutations of a size $n$
are specified by words of length $n$ on the alphabet $[n]$ and without
multiple occurrence of a letter, so that all above definitions about
words remain valid on permutations.
For any nonnegative integer $n$, we write $\nearrow_{n}$
(resp. $\searrow_{n}$)
for the permutation $1 2 \ldots n$ (resp. $n (n-1) \ldots 1$).
If $\pi$ is a permutation of $S_n$,
we denote by $\bar \pi$ the {\em complementary} of $\pi$ that is the
permutation satisfying $\bar \pi(i) = n - \pi(i) + 1$ for all
$i \in [n]$. The {\em inverse} of $\pi$ is denoted by $\pi^{-1}$.

>>>>

Given three words $u$, $v$, and $w$, $u$ is said to a \emph{shuffle} of
$v$ and $w$ if it can be formed by interleaving the letters from $v$ and
$w$ in a way that maintains the left-to-right ordering of the letters
from each word. In addition to purely combinatorial questions, the
shuffle product of words leads to the following purely computational
questions:
\begin{enumerate}
    \item \label{item:problem_1}
    Given three words $u$, $v$, and $w$, decide if $u$ is a shuffle of
    $v$ and $w$.
    \item \label{item:problem_2}
    Given a word $u$, decide if there exist two words $v$ and $w$ such
    that $u$ is a shuffle of $v$ and $w$.
\end{enumerate}
Problem~\ref{item:problem_1} is in $\Pclass$ (it is a classical textbook
exercise to design an efficient dynamic programming algorithm that
determines whether $u$ is a shuffle of $v$ and $w$), while
Problem~\ref{item:problem_2} is in $\NPclass$ as shown
in~\cite{Rizzi:Vialette:CSR:2013}. Analogous questions can be stated to
decide if a word $u$ is a \emph{square} (w.r.t. the shuffle). A square
is a word $u$ with the property that there exists a word $v$ such that
$u$ is a shuffle of $v$ with itself.
\smallskip

This paper is intended to study a natural generalization of the shuffle
to permutations. Roughly speaking, given three permutations $\pi$, $\mu$,
and $\nu$, $\pi$ is said to be a shuffle of $\mu$ and $\nu$ if $\pi$
(viewed as a word) is a shuffle of two words that are order-isomorphic
to $\mu$ and $\nu$.
%For
%example, $\pi$ is the shuffle of $\mu=$ and $\nu=$ as $\pi$. A
%permutation $\pi$ is said to be a square (w.r.t. the shuffle) if there
%exists a permutation $\mu$ such that $\pi$ is a shuffle of $\mu$ with
%itself. For example, $\pi = 241356$ is a square as it is the shuffle of
%$213$ with itself (\emph{e.g.},
%$\pi = \dot{2}{4}\dot{1}{3}{5}\dot{6}$, where the $\dot{x}$'s
%denote the second occurrence), whereas $\pi = 4123$ is not.
Our intention in this paper is to to study this shuffle product of
permutations both from a combinatorial and from a computational point of
view by focusing on {\em square} permutations, that are permutations
$\pi$ being in the shuffle of a permutation $\mu$ with itself. The
shuffle product defined and considered in this paper is different from
other existing and well-known shuffle products on permutations. For
instance, in~\cite{DHT:IJAC:2002}, the authors define the
{\em shifted shuffle product}. In this product $\pi$ is a shuffle of
$\mu$ and $\nu$ if $\pi$ is the shuffle, as words, of $\mu$ and the
word obtained by incrementing all letters of $\nu$ by the size of $\mu$.
It is a simple exercise to prove that, given three permutations $\pi$,
$\mu$, and $\nu$, deciding if $\pi$ is a in the shifted shuffle of
$\mu$ and $\nu$ is in $\Pclass$.
\smallskip

Let us now  briefly review the key results for words that arise in our context.
%Formally, the shuffle of two words $u$ and $v$ of $A^*$ is the language
%$u \shuffle v$ consisting of all words $u_1 v_1 u_2 v_2 \ldots u_k v_k$,
%where $k \geq 0$ and the $u_i$ and the $v_i$ are the words  of $A^*$
%such that $u = u_1 u_2 \ldots u_k$ and $v = v_1 v_2 \ldots v_k$. We
%first briefly review the key results for words that arise in our context.
Given three words $u$, $v_1$ and $v_2$, it can be tested in
$O\left(|u|^2 / \log(|u|)\right)$ time whether or not
$u$ is in $v_1 \shuffle v_2$~\cite{Leeuwen:Nivat:IPL:1982}.
To the best of our knowledge, the first $O(|u|^2)$ time algorithm
appeared in~\cite{Mansfield:DAM:1983}. This algorithm can easily be
extended to check in polynomial-time whether or not a word is the
shuffle of any fixed number of given words. The shuffle $u \shuffle v$
of words $u$ and $v$ can be computed in
\begin{equation}
    O\left((|u|+|v|) \; \binom{|u|+|v|}{|u|}\right)
\end{equation}
time~\cite{Spehner:TCS:1986}.
An improvement and generalization has been proposed in
\cite{Allauzen:IGM:2000}, where it is proved that, given words
$u_1, u_2, \ldots, u_k$, the shuffle $\shuffle_{i=1}^{k} u_i$ can be
computed in
\begin{equation}
    O\left(\binom{|u_1|+|u_2|+\ldots+|u_k|}{|u_1|,|u_2|,\ldots,|u_k|}\right)
\end{equation}
time. Given words $u, v_1, v_2, \ldots, v_n \in A^*$, it is however \NPC
to decide whether or not $u \in \shuffle_{i=1}^{k} v_i$.
\cite{Mansfield:DAM:1983,Warmuth:Haussler:JCSS:1984}.
This remains true even if the alphabet has size $3$~\cite{Warmuth:Haussler:JCSS:1984}.
Of particular interest,
it is shown in~\cite{Warmuth:Haussler:JCSS:1984} that this problem remains
\NPC even if all
words $v_1, v_2, \ldots, v_n$ are identical,
thereby proving that,
for two words $u$ and $v$, it is \NPC to decide whether or not $u$
is in the iterated shuffle of $v$.
Again, this remains true even if the alphabet has size $3$.
As for squares, given a word $z$ it is \NPC to determine whether
there exists a word $x$ such that $z$ is the shuffle of $x$ with itself
\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013}.
\smallskip

This paper is organized as follows...

@@

We denote by $S_n$ the set of permutations of size $n$ and by $S$ the
set of all permutations. In this paper, permutations of a size $n$
are specified by words of length $n$ on the alphabet $[n]$ and without
multiple occurrence of a letter, so that all above definitions about
words remain valid on permutations. The only difference lies on the
fact that we shall denote by $\pi(i)$ (instead of $\pi_i$) the
$i$-th letter of any permutation $\pi$.
For any nonnegative integer $n$, we write $\nearrow_{n}$
(resp. $\searrow_{n}$)
for the permutation $1 2 \ldots n$ (resp. $n (n-1) \ldots 1$).
If $\pi$ is a permutation of $S_n$,
we denote by $\bar \pi$ the {\em complementary} of $\pi$ that is the
permutation satisfying $\bar \pi(i) = n - \pi(i) + 1$ for all
$i \in [n]$. The {\em inverse} of $\pi$ is denoted by $\pi^{-1}$.
